---
title: "Presentation Slides Notebook"
output:
  html_document:
    df_print: paged
---

# Point Patterns

## Objectives
1. Tignan yung events ay mag exhibit ng systematic pattern as oppose to being randomly distributed. Ito ba ay may clustering or regularity sa isang area kesa random or indenpendent scattering.
2. Tignan yung intensity ng mga point pattern kung nag va vary.
3. Tignan ang presenya ng spatial dependence among the events. Ito ay kung asan yung event B ay present kung meron event A.
4. Gumamit ng models para sa mga na observed na patterns.

## Analysis Approach
1. Patterns in event location are the focus.
2. Events may have attributes which can be used to distinguish types.
3. Stochastic aspect is where events are likely to occur.
4. Does a pattern exhibit clustering or regularity?
5. Over what spatial scales do patterns exist?

### 1st Order
* Sa 1st order effects ay interesado lamang tayo sa **means** and **variances**.
* Sa point pattern walang means kundi **intensity**. Tinitignan nito ang mean number of events per unit area. Ang intensity ay essentially a mean.
* Ang mathetical formula ng lambda(s) ay complicated. Ang intensity ay tinitignan yung mean number of event given a location and if you keep on making the area smaller you derive to the intensity. 
* Kung ang mga intensity may maliit lamang na variations, puwede natin masabi na ang point patterns ay **homogeneous**. 

### 2nd Order
* Ang 2nd order property ay ang number of events in pairs of areas.
* Ito ay mga spatial dependence kung saan kung may relasyon ang mga points sa isa't isa -- repel or cluster?
* Mahalaga ito makuha para malaman kung CSR, clustering or regularity ang point patterns dahil may kanya kanyang analysis ang bawat isa.
* Ito ay function ng dalawang location -- product of the two events i and j divided by the area of i and j

## Visualization Issues
1. May underlying population distribution from which events arise in a region.
2. If the population varies we would expect events to cluster in areas of high population. May clustering ba on the basis on the population alone?
3. Are they more or less clustered than we would expect on the basis of population alone.

## Exploring Point Patterns
1st order effects

* Quadrat Methods
* Kernel estimation

2nd order effects

* Nearest neighbors distances
* K-function

### Quadrat methods
Quadrat methods, summarize number of events in each quadrat in region R.
To get intesity, divide the quadrat result in the quadrat area.

Disadvantages:

* Loses spatial detail. Just count the number of events in each quadrat but does not determine if the scatter of the events. Sa modelling dapat ma describe pa rin natin ang events in terms of nearness or scatter.
* Converts data to area value.
* Making quadrats smaller generates *higher variation* and to some extremes some quadrats will have no data or 0.

Advantages:

```{r}
library(spatstat)
data(swedishpines)
x<-swedishpines
Q <- quadratcount(x, nx=4,ny=4)
plot(Q)
plot(round(Q/600,4))
Q2 <- quadratcount(x, nx=8,ny=8)
plot(Q2)
```

### Kernel Estimation
Ang Kernel estimation ay parang moving window Quadrat method. Gumagamit ito ng kernel at edge correction. Ang kernel ay ang moving window at posibleng may window kung saan kasali ang exterior ng study region. Ito ang tinatawag na edge correction.

Ang tau ay smoothing parameter.Ginagamit ito sa density mapping para ma visualize yung intensity. Imbes na moving window, ang kernel ay bilog at may radius, at kikuha lahat ng events sa loob.

Kapag malaki ang tau mas smooth, pero pag maliit may spikes on may improve yung intensity.

The contribution of an event can extend in multiple kernels unlike the contribution of event in quadrat in which is only in 1 quadrat.

**Adaptive Kernel Estimation** is one way to correct edge defects. Pwede iba iba ang tau dito kesa sa constant tau.
```{r}
plot(density(x,5),axes=FALSE)
plot(density(x,10),axes=FALSE)
plot(density(x,15),axes=FALSE)
```

### Nearest neighbor distances
Ito ay ginagamit upang mag investigate 2nd oder properties - kung ang pattern ba ay CSR, clustering or may regularity. Ang mga tanong ay, kung ang distribution ng distances. Kung may maraming nearest neighbor na maliit ang distances, pwede masabi na may clustering at kung malalaki naman ang distances ay maaring masabi na may regularity.

* event - event, distance between a random event to its nearest event, it can be a tall tree near a tall tree, both are events, this is actual data
* point - event, distance between a selected point to an event, a point could be empty, given a person location, is there a tall tree? this is representation of randomness,

Given ng dataset, ang W ay nearest neighbor of each event. This is the statistical measure, and will get a collection of metrics.
$$G_{hat}(w)=\frac{w_{no. events} \leq w_{nominated}}{n}$$
Ang G function ay gamit sa event-event (w), at F function ay sa point-event (x)

* Ang early sharpening could indicate clustering since marami na may close Ghat sa lower values ng w(distance). Dominant sa smaller w values.
* Ang late sharpening naman indicated repulsion kasi sa higher values of w(distance) na nagkaroon ng maraming distribution/frequency. Dominant sa high w values.

Dahil sa mga graph, ang CSR ay nag eexhibit ng stair-like juggedness. Or in general dapat sloping upwards.

EDA is exploratory data analysis.
```{r}
Fc <- Fest(cells) 
Gc <- Gest(cells) 
Kc <- Kest(cells) 
Lc <- Lest(cells) 

plot(Fc,rs~r,main="Border-corrected Estimate of F")
plot(Fc,theo~r,main="Poisson Estimate of F")
plot(Fc)
plot(Gc,rs~r,main="Border-corrected Estimate of G")
plot(Gc,theo~r,main="Poisson Estimate of G")
plot(Gc)
plot(Kc,border~r,main="Border-corrected Estimate of K")
plot(Kc,theo~r,main="Poisson Estimate of K")
plot(Kc)
plot(Lc,border~r,main="Border-corrected Estimate of L")
plot(Lc,theo~r,main="Poisson Estimate of L")
plot(Lc)
```

What if nearest events are outside study region? and are clustered on one corner? This introduce a correction. Ang edge correction ay nag dedefine ng exterior bound. Yung mga nasa labas ay hindi hahanapan ng nearest neighbor pero yung events sa loob neto makaka contribute sa analysis ng study region.

### K-function

* Ang limitation ng nearest neighbors considers only shortest scales.
* Given a location, ang intensity at relationship with other events are regardless of direction.
* Isotrophy, direction does not matter.
* It compares the expected events to real events.


## Modeling Spatial Point Patterns

### Objectives
* To test a baseline hypothesis
* To construct models to explain observed patterns
* The standard model against which a spatial point pattern is compared is **Complete Spatial Randomness**

### Complete Spatial Randomness (CSR)
Reasons for beginning an analysis with a test for CSR

* Rejection of CSR is prerequisity for attempt to model an observed pattern.
* Test can assist to explore a datase and assist in the formulation of alternatives to CSR.
* CSR operates as dividing hypothesis between regular and clustered patterns.

The hypothesis asserts that:

1. The number of events in the study region follows a Poisson distribution. This implies constant intensity, no 1st order effects.
2. Given *n* events in *A*, the events are an independent ramdom sample from a uniform distribution on A. This implis no spatial interaction.

Remarks:
1. A pattern for which CSR is not rejected merits further formal statistical analysis.


## Test for CSR
### Simple Quadrat Tests for CSR

$$X^2=\frac{(m-1)S^2}{\bar X}$$
* S^2 = sample variance, Xbar = sample mean
* When statistics is large = clustering, small = regularity.
* Sa R, p-values na ang ipina pakita, X^2 is the p-value

$$ICS=\frac{S^2}{\bar X}-1$$
* ICS < 0, suggestion of regularity,
* ICS > 0, clustering
* ICS = 0, CSR
* Disadvatge: unequal quadrat size affect analysis

```{r}
data("swedishpines")
M <- quadrat.test(swedishpines,nx=3,ny=4)
M
plot(M) 
plot(swedishpines)
plot(M, add=T, cex=1)
```



### Nearest Neighbor Tests for CSR
```{r}
  ## without edge correction ##
clarkevans.test(swedishpines) 

## with guard area (edge correction) ##
clip1<-owin(c(20,50),c(20,50))
clarkevans.test(swedishpines,correction="guard",clipregion=clip1) 
clip2<-owin(c(20,80),c(20,80))
clarkevans.test(swedishpines,correction="guard",clipregion=clip2) 
```
### K-function Tests for CSR




## Analysis of Multiple Types of Events
* A multivariate spatial point process where event have classification types.
* The univariate are referred as components of multivariate process.

Analysis Objectives

* Detection of relationship in the pattern of one type from another.
* Identify indepence of types of events as opposed to attraction or repulsion.

* One event may attract or repel another event. CSR cannot be connected with multivariate, but for individual process.
* Hypothesizes a positive or negative dependence among the event types.
* Independence does not imply that any of the component need to be CSR.

## How to analyze multivariate data

### Quadrat Count Analysis
This works by using cross tabulations. Kada quarant ay may count ng present at absent ng types. Compare to 3.84, kung malaki ay wala silang reason to be independent.
```{r}
data("lansing")
plot(lansing,main="Lansing Woods")
plot(split(lansing))
plot(density(split(lansing)),ribbon=F)
plot(Kcross(lansing,"redoak","whiteoak"))
plot(Kdot(lansing,"hickory"))

data(amacrine)
plot(amacrine,main="Rabbit Amacrine Cells")
Env <- envelope(amacrine,Kcross,nsim=99,i="on",j="off")
plot(Env)
```

### Quadrat Count Analysis
* G(h) Given event i, whats the nearest even j, the probability of getting another type of event?
* F(h) given a point, nearest event of j is less that or equal to h
* For an independent patter: G(h) = F(h), the nearest distances from events of j to randomly selected event of point must be the same.

### Bivariate or Coss K-Function
* Under assumed independence, type i events should be random with respect to type j events.



## Space-Time Interaction
Given events that are close in time (nangyari), are there events in space closer than would be expected, mas sabog ba earlier? Observe where true events happen as time goes by. Spatiotemporal.

Example: Onset of disease on days.

* The CSTR (Complete Spatiotemporal Randomness) is the absence of structure in time as well in space.
* Space-time clustering is the alternative to CSTR
* Space time clustering is said to exist if, among those events that are closer in time, there are events that are closer in space.


### Knox test for space-time interaction
* Quantifies based on critical space and time distances.
* Uses s and tau, for critical space and time distances respectively

$$X=\sum_{i=1} ^N \sum_{j=1} ^{i-1}s_{ij}t_{ij}$$
### Mantel test for space-time interaction
* This avoids problem of determining critical distances, rather get the average space and time distances
* Mantel is more popular than Knox. It also implies correlation.

```{r}
Input =("
Onset Lat Lon
        72	34.13583	117.9236
        59	34.17611 	118.3153
        41	33.82361 	118.1875
        72	34.19944 	118.5347
        61	34.06694 	117.7514
        32	33.92917 	118.2097
        52	34.01500 	118.0597
        47	34.06722 	118.2264
        65	34.08333 	118.1069
        75	34.38750 	118.5347
        ")

MantelData = read.table(textConnection(Input),header=TRUE)
space.dists <- dist(cbind(MantelData$Lon, MantelData$Lat))
time.dists <- dist(MantelData$Onset)
as.matrix(space.dists)
as.matrix(time.dists)

install.packages("ade4") 
library(ade4)
mantel.rtest(space.dists, time.dists, nrepet = 9999)
```
### K-function for Space-time interaction
* If there is space time interaction, D(h,t) will be large


## Clustering around a Point Source


### Neyman-Scott process
```{r}
library(spatstat)

data("redwood")
x <- redwood
plot(x, main = "Strauss-Ripley Redwood Saplings Dot Map")
summary(x)

qa <- quadratcount(x, 25,25)
plot(intensity(qa, image = TRUE, main = "25x25 Quadrat Map of Redwood Saplings"))


#Viasualization of Intensity using Kernel Estimation

plot(density(x, 0.05), axes = TRUE, main = "Bandwith = 0.05")
plot(x, add = TRUE)
contour(density(x, 0.05), axes = FALSE, add = TRUE)

plot(density(x, 0.10), axes = TRUE, main = "Bandwidth = 0.10")
plot(x, add = TRUE)
contour(density(x, 0.10), axes = FALSE, add = TRUE)

#Test for CSR
#Quadrat Test

QT <- quadrat.test(x,3) #two-sided test on a 3x3 grid
QT
plot(QT, main = "Quadrat Test on a 3X3 grid")
plot(x, add = TRUE)
quadrat.test(x, 3, alternative = "clustered") #one-sided test for clustering

plot(intensity(quadratcount(x,3), image = TRUE), main = "3x3 Quadrat Map Saplings")

#Clark-Evans Test

clarkevans.test(x) #no edge correction and two-sided test
clarkevans.test(x, alternative = "clustered")
#no edge correction and one-sided test for clustering

clip<- owin(c(0.2, 0.8),c(-0.8,-0.2)) #defines the "desired interior" for edge correction
clarkevans.test(x, correction = "guard", clipregion = clip)
#with edge correction and two-sided test

clarkevans.test(x, correction = "guard", clipregion = clip, alternative = "clustered")
#with edge correction and one-sided test for clustering

#Nearest-Neighbor Distance and Simulation Envelopes
Fc <- Fest(x)
Gc <- Gest(x)
Lc <- Lest(x)
Kc <- Kest(x)

plot(Fc)
Fenv <- envelope(x, Fest, nrank = 1, nsim = 999)
plot(Fenv, main = "Simulation Envelope for F")

plot(Gc)
Genv <- envelope(x, Gest, nrank=1, nsim = 999)
plot(Genv, main = "Simulation Envelopes for G")

plot(Lc)
Lenv <- envelope(x, Lest, nsim= 999, rank = 1)
plot(Lenv, main = "Simulation Envelopes for L")

plot(Kc)
Kenv <- envelope(x, Kest, nsim = 999, rank =1)
plot(Kenv, main = "Simulation Envelopes for K")

#Model Fitting under Neyman-Scott Process

model1 <- kppm(x, trend = ~1, clusters = "Thomas", statistic = "K")
plot(model1)
summary(model1)

model1Kenv <- envelope(model1, Kest, nsim = 999, rank =1)
plot(model1Kenv)
model1Lenv <- envelope(model1, Lest, nsim = 999, rank =1)
plot(model1Lenv)

#KPPM

kppm(x,
     trend = ~1,
     clusters = c("Thomas", "MatClust","Cauchy","VarGamma","LGCP"),
     method = c("mincon", "clik2", "palm"),
     statistic = c("K","pcf"),
     )
```

